Retrofitting Effect Handlers onto OCaml

### 2020-11-07 retro-concurrency.tex
Updates

In this section, we measure the impact of the addition of effect handlers on
code that does not use effect handlers. Our macro benchmark suite consists of
54 real OCaml workloads including verification tools (|Coq|, |Cubicle|,
|AltErgo|), parsers (|menhir|,|yojson|), storage engines (|irmin|), utilities
(|cpdf|,|decompress|), bioinformatics (|fasta|, |knucleotide|, |revcomp2|,
|regexredux2|), numerical analysis (|grammatrix|,|LU_decomposition|) and
simulations (|nbody|,|game_of_life|). In addition to Stock (|stock|) and
Multicore OCaml (|MC|), we also ran the benchmarks on Multicore OCaml with no
red zone (|MC+RedZone0|) in the fiber (all OCaml functions will have stack
overflow check) and a red zone size of 32 words (|MC+RedZone32|). Recall that
the default red zone size in Multicore OCaml is 16 words.

@@

Next we look at the cost of setting up an effect handler, but not |perform|ing
an effect. The aim is to quantify the overhead of setting up and tearing down
effect handlers compared to a non-tail function call. To this end, we surround
the non-tail calls in the recursive micro benchmarks with an effect handler. We
also include a version of the same benchmarks implemented using concurrency
monad~\cite{} (|monad|) as a proxy for CPS versions. Recall that OCaml compiler
does not use CPS in its IR. In the |monad| version, we use a |fork| to invoke
the non-tail call and use an |MVar| to collect its result.

@@

Next we look at benchmarks that utilize non-local control flow. First, we
quantify the overhead individual operations in effect handling. Consider the
following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took \textbf{48ns,
10ns, 13ns and 19ns}, respectively. For reference, the idle memory latency for
the local NUMA node is 93.2ns as measured using the Intel MLC tool~\cite{}.

\paragraph{Generators} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers, given any data structure (|'a t|)
and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can derive its
generator function (|val next : unit -> 'a option|) (|MC| version). The code is
available in the supplementary material.

We evaluate the performance of traversing a complete binary tree of depth 25
using this generator. For comparison, we also implemented a hand-written,
selective CPSed~\cite{}, defunctionalised~\cite{DefunctionalizationAtWork}
version (|cps|) and a concurrency monad (|monad|) version of the generator for
the tree. Both |cps| and |monad| versions are specialised to the binary tree,
and come with the usual caveats of not using the stack for function calls. We
observed that the |cps| version was the fastest thanks to specialisation and
hand optimisation. |MC| version was only 2.76$\times$ slower than |cps| while
being a generic solution, whereas the |monad| version was 8.69$\times$ slower
than |cps|.

@@

\subsection{Effects benchmarks}

We now look at the impact of the overhead in a variety of benchmarks comparing
the cost of using our effects abstraction versus alternative methods of
implementation.

TODO:
\begin{itemize}
\item Aim: |R3| Do we have an efficient implementation of effects and are effects competitive vs alternative implementation strategies.
\item we can show the sequential vs stack benchmarks as used by 'Folklore to fact' \tk{slightly weak story as the sequential ones will be much faster for these very simple benchmarks and effects don't really add anything to their implementation}
\item general generator examples (using examples from effects-example)
\item chamenos vs chamenos effects vs lwt (using examples from effects-example)
\end{itemize}

>>>>

In this section, we measure the impact of the addition of effect handlers on
code that does not use effect handlers. Our macro benchmark suite consists of
54 real OCaml workloads including verification tools (|Coq|, |Cubicle|,
|AltErgo|), parsers (|menhir|, |yojson|), storage engines (|irmin|), utilities
(|cpdf|, |decompress|), bioinformatics (|fasta|, |knucleotide|, |revcomp2|,
|regexredux2|), numerical analysis (|grammatrix|, |LU_decomposition|) and
simulations (|nbody|, |game_of_life|). In addition to Stock (|stock|) and
Multicore OCaml (|MC|), we also ran the benchmarks on Multicore OCaml with no
red zone (|MC+RedZone0|) in the fiber (all OCaml functions will have stack
overflow check) and a red zone size of 32 words (|MC+RedZone32|). Recall that
the default red zone size in Multicore OCaml is 16 words.

@@

Next, we aim to quantify the overhead of setting up and tearing down effect
handlers compared to a non-tail function call. To this end, we surround the
non-tail calls in the recursive micro benchmarks with an effect handler. The
code does not |perform| effects. We also include a version of the same
benchmarks implemented using concurrency monad~\cite{} (|monad|) as a proxy for
CPS versions. Recall that OCaml compiler does not use CPS in its IR. In the
|monad| version, we use a |fork| to invoke the non-tail call and use an |MVar|
to collect its result.

@@

Next we look at benchmarks that utilize non-local control flow using effect
handlers. First, we quantify the overhead individual operations in effect
handling. Consider the following annotated code:
\begin{lstlisting}
effect E : unit
(* a *) match (* b *) perform E (* d *) with
| v -> (* e *) v
| effect E k -> (* c *) continue k ()
\end{lstlisting}
The sequence \texttt{a-b} involves allocating a new fiber and switching to it.
\texttt{b-c} is performing the effect and handling it. \texttt{c-d} is resuming
the continuation. \texttt{d-e} is returning from the fiber with a value and
freeing the fiber. We measured the time taken to execute these sequences using
|perf| support for cycle-accurate tracing on modern Intel processors. We
executed 10 iterations of the code, with 3 warm-up runs. We observed that the
sequences \texttt{a-b}, \texttt{b-c}, \texttt{c-d} and \texttt{d-e} took
\textbf{48ns, 10ns, 13ns and 19ns}, respectively. For reference, the idle
memory latency for the local NUMA node is 93.2ns as measured using the Intel
MLC tool~\cite{}.

\paragraph{Generators} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{},
defunctionalised~\cite{DefunctionalizationAtWork} version (|cps|) and a
concurrency monad (|monad|) version of the generator for the tree. Both |cps|
and |monad| versions are specialised to the binary tree, and come with the
usual caveats of not using the stack for function calls. We observed that the
|cps| version was the fastest thanks to specialisation and hand optimisation.
|MC| version was only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.

\paragraph{Chameneos} Chameneos~\cite{} is a concurrency game aimed at
measuring context switching and synchronization overheads. Our implementation
uses MVars~\cite{} for synchronization. We compare effect handler (|MC|),
concurrency monad (|monad|) and Lwt, a widely used concurrency programming
library for OCaml (|lwt|) versions. We observed that |MC| was the fastest, and
|monad| (|lwt|) was 1.67$\times$ (4.29$\times$) slower than |MC|.

@@
